---
title: "Data Description and Exploratory Analysis (Rossmann)"
author: "Kayra Işık"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 8, fig.height = 4.5)
set.seed(42)
```

# Purpose

A full description of the Rossmann data for the Data chapter of the thesis: the
variables, descriptive statistics of the target, the within-week and
within-year structure, promotion and store-type effects, an individual-store
view, and a check of the observed properties against the assumptions behind the
decomposition model. Every figure is written to `eda_outputs/` as a PDF so it
can be dropped straight into the LaTeX Data section, and the key tables are
printed here for the descriptive-statistics and feature tables.

```{r packages}
library(tidyverse)
library(lubridate)
BLUE <- "#4C72B0"; RED <- "#C44E52"; GREY <- "grey40"
theme_set(theme_minimal(base_size = 12))
dir.create("eda_outputs", showWarnings = FALSE)
save_fig <- function(p, name, w = 5, h = 4)
  ggsave(file.path("eda_outputs", name), p, width = w, height = h)
```

# 1. Raw to final modelling sample

```{r load}
train_raw <- read_csv("train.csv", show_col_types = FALSE) %>% mutate(Date = as.Date(Date))
store_raw <- read_csv("store.csv", show_col_types = FALSE)

steps <- tibble(
  Step = c("Raw train.csv rows",
           "After removing closed days (Open = 0)",
           "After also removing zero-sales open days",
           "Final modelling sample"),
  Rows = c(nrow(train_raw),
           nrow(filter(train_raw, Open == 1)),
           nrow(filter(train_raw, Open == 1, Sales > 0)),
           nrow(filter(train_raw, Open == 1, Sales > 0))))
steps
```

```{r build}
df <- train_raw %>% filter(Open == 1, Sales > 0) %>%
  left_join(store_raw, by = "Store") %>%
  mutate(Month     = month(Date, label = TRUE),
         MonthNum  = month(Date),
         YearMonth = floor_date(Date, "month"),
         DOW       = factor(DayOfWeek, levels = 1:7,
                            labels = c("Mon","Tue","Wed","Thu","Fri","Sat","Sun")),
         StoreType = factor(StoreType, levels = c("a","b","c","d"),
                            labels = c("A","B","C","D")),
         Promo     = factor(Promo, levels = c(0,1), labels = c("No promo","Promo")))
cat("Final sample:", nrow(df), "open store-days,", n_distinct(df$Store), "stores,",
    as.character(min(df$Date)), "to", as.character(max(df$Date)), "\n")
```

# 2. Descriptive statistics of the target

```{r desc-stats}
sk <- function(x){ x <- x[is.finite(x)]; m <- mean(x); s <- sd(x); mean(((x-m)/s)^3) }
desc <- tibble(
  Statistic = c("Mean","Median","Std Dev","Min","Max","Skewness"),
  Value = c(mean(df$Sales), median(df$Sales), sd(df$Sales),
            min(df$Sales), max(df$Sales), sk(df$Sales)))
desc %>% mutate(Value = round(Value, 2))
```

```{r fig1-distribution}
p_raw <- ggplot(df, aes(Sales)) +
  geom_histogram(bins = 60, fill = BLUE) +
  labs(x = "Daily sales (EUR)", y = "Count", title = "Raw")
p_log <- ggplot(df, aes(log(Sales))) +
  geom_histogram(bins = 60, fill = BLUE) +
  labs(x = "log(daily sales)", y = "Count", title = "Log-transformed")
print(p_raw); print(p_log)
save_fig(p_raw, "fig_eda_dist_raw.pdf"); save_fig(p_log, "fig_eda_dist_log.pdf")
```

# 3. Trend and seasonality

```{r fig2-over-time, fig.height=4}
p_time <- df %>% group_by(YearMonth) %>%
  summarise(mean_sales = mean(Sales), .groups = "drop") %>%
  ggplot(aes(YearMonth, mean_sales)) +
  geom_line(colour = BLUE, linewidth = 1) + geom_point(colour = BLUE) +
  labs(x = NULL, y = "Mean daily sales (EUR)",
       title = "Average daily sales over time (monthly)")
print(p_time); save_fig(p_time, "fig_eda_sales_over_time.pdf", w = 8, h = 4)
```

```{r fig3-seasonality}
p_dow <- df %>% group_by(DOW) %>% summarise(m = mean(Sales), .groups="drop") %>%
  ggplot(aes(DOW, m)) + geom_col(fill = BLUE) +
  labs(x = NULL, y = "Mean sales (EUR)", title = "By day of week")
p_mon <- df %>% group_by(Month) %>% summarise(m = mean(Sales), .groups="drop") %>%
  ggplot(aes(Month, m, group = 1)) + geom_line(colour = BLUE, linewidth = 1) +
  geom_point(colour = BLUE) + labs(x = NULL, y = NULL, title = "By month")
print(p_dow); print(p_mon)
save_fig(p_dow, "fig_eda_dow.pdf"); save_fig(p_mon, "fig_eda_month.pdf")
```

# 4. Promotions and store types

```{r fig4-promo-type}
p_promo <- df %>% group_by(Promo) %>% summarise(m = mean(Sales), .groups="drop") %>%
  ggplot(aes(Promo, m, fill = Promo)) + geom_col() +
  scale_fill_manual(values = c(GREY, RED)) +
  labs(x = NULL, y = "Mean sales (EUR)", title = "Promotion effect") +
  theme(legend.position = "none")
p_type <- df %>% group_by(StoreType) %>% summarise(m = mean(Sales), .groups="drop") %>%
  ggplot(aes(StoreType, m, fill = StoreType)) + geom_col() +
  labs(x = "Store type", y = NULL, title = "By store type") +
  theme(legend.position = "none")
print(p_promo); print(p_type)
save_fig(p_promo, "fig_eda_promo.pdf"); save_fig(p_type, "fig_eda_storetype.pdf")
```

```{r fig-violin, fig.height=4}
p_violin <- ggplot(df, aes(StoreType, Sales, fill = StoreType)) +
  geom_violin(alpha = .7) + geom_boxplot(width = .12, outlier.size = .2, fill = "white") +
  labs(x = "Store type", y = "Daily sales (EUR)",
       title = "Sales distribution by store type") +
  theme(legend.position = "none")
print(p_violin); save_fig(p_violin, "fig_eda_storetype_violin.pdf", w = 7, h = 4)
```

# 5. An individual store

```{r fig5-store}
set.seed(1); s_id <- 1
s1 <- df %>% filter(Store == s_id) %>% arrange(Date)
p_hist <- ggplot(s1, aes(Date, Sales)) + geom_line(colour = BLUE, linewidth = .3) +
  labs(x = NULL, y = "Daily sales (EUR)", title = paste0("Store ", s_id, ": full history"))
p_spromo <- s1 %>% group_by(Promo) %>% summarise(m = mean(Sales), .groups="drop") %>%
  ggplot(aes(Promo, m, fill = Promo)) + geom_col() +
  scale_fill_manual(values = c(GREY, RED)) +
  labs(x = NULL, y = "Mean sales (EUR)", title = paste0("Store ", s_id, ": promotion effect")) +
  theme(legend.position = "none")
print(p_hist); print(p_spromo)
save_fig(p_hist, "fig_eda_store_history.pdf", w = 7, h = 3.5)
save_fig(p_spromo, "fig_eda_store_promo.pdf", w = 4, h = 3.5)
```

```{r fig-examples-by-type, fig.height=5}
samp <- df %>% distinct(Store, StoreType) %>% group_by(StoreType) %>%
  slice_sample(n = 1) %>% pull(Store)
p_ex <- df %>% filter(Store %in% samp) %>%
  ggplot(aes(Date, Sales)) + geom_line(linewidth = .25, colour = BLUE) +
  facet_wrap(~ paste0("Store ", Store, " (Type ", StoreType, ")"),
             ncol = 1, scales = "free_y") +
  labs(x = NULL, y = "Daily sales (EUR)",
       title = "Example store series, one per store type")
print(p_ex); save_fig(p_ex, "fig_eda_example_series.pdf", w = 9, h = 5)
```

# 6. Feature relationships

```{r fig-competition, fig.height=4}
p_comp <- df %>% filter(!is.na(CompetitionDistance)) %>%
  mutate(dist_bin = cut(CompetitionDistance,
                        breaks = quantile(CompetitionDistance, 0:10/10, na.rm = TRUE),
                        include.lowest = TRUE)) %>%
  group_by(dist_bin) %>%
  summarise(m = mean(Sales), mid = median(CompetitionDistance), .groups = "drop") %>%
  ggplot(aes(mid, m)) + geom_line(colour = BLUE) + geom_point(colour = BLUE) +
  labs(x = "Competitor distance (m, decile midpoints)", y = "Mean sales (EUR)",
       title = "Sales vs. competitor distance")
print(p_comp); save_fig(p_comp, "fig_eda_competition.pdf", w = 7, h = 4)
```

```{r fig-corr, fig.height=5}
num <- df %>% transmute(Sales, DayOfWeek, Promo = as.integer(Promo) - 1,
                        SchoolHoliday, CompetitionDistance,
                        Promo2, Month = MonthNum) %>%
  mutate(across(everything(), as.numeric)) %>% drop_na()
cc <- cor(num)
p_corr <- as_tibble(cc, rownames = "x") %>%
  pivot_longer(-x, names_to = "y", values_to = "r") %>%
  ggplot(aes(x, y, fill = r)) + geom_tile() +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3) +
  scale_fill_gradient2(low = RED, mid = "white", high = BLUE, midpoint = 0, limits = c(-1,1)) +
  labs(x = NULL, y = NULL, title = "Correlations among numeric features") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p_corr); save_fig(p_corr, "fig_eda_corr.pdf", w = 6.5, h = 5.5)
```

# 7. Store types summary table

```{r storetype-table}
v_by_store <- df %>% group_by(Store) %>%
  summarise(v = var(Sales),
            v_dow = var(ave(Sales, DOW)),
            .groups = "drop") %>%
  mutate(weekly_share = v_dow / v)
type_tab <- df %>% group_by(Store, StoreType) %>%
  summarise(mean_sales = mean(Sales), .groups = "drop") %>%
  left_join(v_by_store, by = "Store") %>%
  group_by(StoreType) %>%
  summarise(Stores = n(),
            `Mean open-day sales (EUR)` = round(mean(mean_sales)),
            `Weekly-predictable share` = round(mean(weekly_share, na.rm = TRUE), 2),
            .groups = "drop")
type_tab
```

# 8. Decomposition view (assumptions check)

The model writes sales as a store level plus predictable, systematic, and noise
components, $y_{s,t} = \mu_s + \delta_{s,t} + \phi_{s,t} + \varepsilon_{s,t}$.
An STL decomposition of one store's series shows the trend, the weekly seasonal
piece (a visual of $\delta$), and the remainder, and indicates how stable the
store level $\mu_s$ is over the sample.

To keep the weekly cycle aligned to actual calendar days, the series is rebuilt
on the full daily grid (so a Sunday is always the same position in the cycle),
closed and missing days are filled by linear interpolation rather than dropped,
and the series starts on the first Monday so the seven seasonal indices map to
Monday through Sunday. This avoids the misalignment that arises when an
open-day-only series is treated as if every seven trading days made one week.

```{r fig-stl, fig.height=6}
library(zoo)
# Store 1 has weak weekly structure, so it is a poor illustration of delta.
# Pick an illustrative store with a strong, clear weekly pattern (high share of
# variance explained by day-of-week) among stores with a full history.
stl_pick <- df %>% group_by(Store) %>%
  summarise(dow_share = var(ave(Sales, DOW)) / var(Sales), n = n(), .groups = "drop") %>%
  filter(n > 700) %>% slice_max(dow_share, n = 1)
stl_id <- stl_pick$Store
cat("STL example store:", stl_id, "| weekly-predictable share:",
    round(stl_pick$dow_share, 2), "\n")
s_stl <- df %>% filter(Store == stl_id) %>% arrange(Date)

# full daily calendar for the store, with open-day sales joined in
full <- tibble(Date = seq(min(s_stl$Date), max(s_stl$Date), by = "day")) %>%
  left_join(s_stl %>% select(Date, Sales), by = "Date")
# start on the first Monday so cycle position 1 = Monday (week_start = 1)
first_mon <- full$Date[which(wday(full$Date, week_start = 1) == 1)[1]]
full <- full %>% filter(Date >= first_mon)
# fill closed/missing days: linear interpolation inside, carry-forward/back at edges
sfill <- na.approx(full$Sales, na.rm = FALSE)
sfill <- na.locf(na.locf(sfill, na.rm = FALSE), fromLast = TRUE)

ts1 <- ts(sfill, frequency = 7)               # 7 = true calendar week (Mon..Sun)
# t.window: STL's default (~11 days) makes the trend far too flexible on a
# multi-year daily series, so it absorbs the weekly cycle. A wide window (~26
# weeks) forces a genuinely slow trend and leaves the weekly signal in the
# seasonal term. robust = TRUE downweights promotion/holiday spikes.
dec <- stl(ts1, s.window = "periodic", t.window = 181, robust = TRUE)
pdf("eda_outputs/fig_eda_stl.pdf", width = 8, height = 6)
plot(dec, main = paste("STL decomposition (calendar-aligned), Store", stl_id))
dev.off()
plot(dec, main = paste("STL decomposition (calendar-aligned), Store", stl_id))

# how much of the variance is weekly seasonal vs trend vs remainder
comp <- as.data.frame(dec$time.series)
tibble(seasonal = var(comp$seasonal), trend = var(comp$trend),
       remainder = var(comp$remainder)) %>%
  mutate(across(everything(), ~ round(.x / sum(c(var(comp$seasonal),
         var(comp$trend), var(comp$remainder))) * 100, 1))) %>%
  rename_with(~ paste0(.x, "_var%"))
```

```{r session}
sessionInfo()
```
