---
title: "Simulation Study: A Controlled Test of the Decomposition Framework"
author: "Kayra Işık"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
set.seed(42)
```

# Purpose

The framework predicts that the gain of a tree-based estimator over a
store-mean forecast is approximately
\[
  \mathrm{MSE}^{\text{naive}} - \mathrm{MSE}^{\text{tree}}
  \;\approx\; \mathrm{Var}(\delta)\,\rho^2,
\]
where \(\mathrm{Var}(\delta)\) is the amount of predictable within-store
variation and \(\rho^2\) is how well the available features let the tree recover
it. This file tests that prediction on **simulated data with a known
data-generating process (DGP)**, so the two quantities can be controlled
directly. It is the "running example'' of the thesis: a transparent setting in
which the theoretical mechanism can be switched on and off.

## The data-generating process

For store \(s=1,\dots,N\) and day \(t=1,\dots,T\), sales are generated as
\[
  y_{s,t} = \mu_s + \delta_{s,t} + \phi_{s,t} + \varepsilon_{s,t},
\]
matching the decomposition used in the theoretical framework:

- \(\mu_s\) is a store fixed effect (long-run level), drawn once per store;
- \(\delta_{s,t}\) is the **predictable** component, a function of *observable*
  features (day-of-week and promotion), whose variance is set by `sd_delta`;
- \(\phi_{s,t}\) is a **systematic but unobservable** component, driven by a
  latent factor the estimator never sees, with variance set by `sd_phi`;
- \(\varepsilon_{s,t}\) is mean-zero noise with standard deviation `sd_eps`.

Only day-of-week and promotion enter the estimator's feature matrix, so the
store-mean baseline leaves all of \(\delta\) unexplained, while the tree can
recover the part of \(\delta\) that the features identify. Two knobs are varied:
the strength of the predictable signal (`sd_delta`) and the share of \(\delta\)
the features can actually recover (`signal_share`, a proxy for \(\rho^2\)).

# 1. Packages and helpers

```{r packages}
library(tidyverse)
library(lightgbm)

N_THREADS <- max(1, parallel::detectCores() - 1)

# Generate one panel data set under the DGP.
simulate_panel <- function(N = 150, T = 364, sd_mu = 2000,
                           sd_delta = 600, signal_share = 0.8,
                           sd_phi = 300, sd_eps = 300, seed = 1) {
  set.seed(seed)
  dow_effect   <- rnorm(7, 0, 1)                 # weekly shape (shared)
  promo_effect <- 1.2                            # promotion lift (shared)
  grid <- expand_grid(store = 1:N, t = 1:T) %>%
    mutate(
      dow    = ((t - 1) %% 7) + 1,
      promo  = rbinom(n(), 1, 0.35),
      mu     = rnorm(N, 6000, sd_mu)[store],
      # predictable component from OBSERVABLE features (dow + promo):
      delta_obs = dow_effect[dow] + promo_effect * promo,
      # a predictable-looking part driven by a HIDDEN factor (becomes phi):
      hidden    = rnorm(n()),
      delta_hid = hidden
    )
  # scale the observable and hidden predictable parts so that
  #   Var(delta_obs) / (Var(delta_obs)+Var(delta_hid)) = signal_share,
  #   and total predictable sd = sd_delta.
  vo <- var(grid$delta_obs); vh <- var(grid$delta_hid)
  a  <- sqrt(signal_share)        / sqrt(vo)
  b  <- sqrt(1 - signal_share)    / sqrt(vh)
  grid <- grid %>%
    mutate(
      delta = sd_delta * (a * delta_obs),         # recoverable from features
      phi   = sd_delta * (b * delta_hid) + rnorm(n(), 0, sd_phi),
      eps   = rnorm(n(), 0, sd_eps),
      sales = mu + delta + phi + eps
    )
  grid
}
```

> Note: in the DGP the *recoverable* predictable variance is `Var(delta)` and
> the unrecoverable systematic part is folded into `phi`. The framework's
> predicted gap is therefore `Var(delta)` (the part the features identify),
> which we read directly from the simulated series below.

# 2. Single illustrative panel

```{r single}
df <- simulate_panel(N = 150, T = 364, sd_delta = 600, signal_share = 0.8)
cut <- floor(max(df$t) * 0.85)
tr  <- df %>% filter(t <= cut); te <- df %>% filter(t > cut)
feats <- c("store","dow","promo")
sm <- tr %>% group_by(store) %>% summarise(m = mean(sales), .groups="drop")
mse_naive <- mean((te$sales - (te %>% left_join(sm, by="store") %>% pull(m)))^2)
mm <- lgb.train(list(objective="regression", metric="rmse", num_leaves=31,
                     learning_rate=0.1, num_threads=N_THREADS, verbosity=-1),
                lgb.Dataset(as.matrix(tr[,feats]), label=tr$sales),
                nrounds=200, verbose=-1)
mse_tree <- mean((te$sales - predict(mm, as.matrix(te[,feats])))^2)
cat(sprintf("Empirical gap (MSE_naive - MSE_tree): %.0f\n", mse_naive - mse_tree))
cat(sprintf("Var(delta) (recoverable signal):      %.0f\n", var(df$delta)))
```

# 3. Grid: gap versus Var(delta) and the recoverable share

```{r grid, cache=TRUE}
run_cell <- function(sd_delta, signal_share, rep) {
  d <- simulate_panel(N = 120, T = 364, sd_delta = sd_delta,
                      signal_share = signal_share, seed = 1000*rep + sd_delta)
  cut <- floor(max(d$t) * 0.85); tr <- d %>% filter(t<=cut); te <- d %>% filter(t>cut)
  sm <- tr %>% group_by(store) %>% summarise(m=mean(sales), .groups="drop")
  mn <- mean((te$sales - (te %>% left_join(sm,by="store") %>% pull(m)))^2)
  mo <- lgb.train(list(objective="regression",metric="rmse",num_leaves=31,
                       learning_rate=0.1,num_threads=N_THREADS,verbosity=-1),
                  lgb.Dataset(as.matrix(tr[,c("store","dow","promo")]),label=tr$sales),
                  nrounds=200, verbose=-1)
  mt <- mean((te$sales - predict(mo, as.matrix(te[,c("store","dow","promo")])))^2)
  tibble(sd_delta = sd_delta, signal_share = signal_share,
         var_delta = var(d$delta), gap = mn - mt)
}

grid <- expand_grid(sd_delta = c(200, 400, 600, 800),
                    signal_share = c(0.3, 0.6, 0.9), rep = 1:3) %>%
  pmap_dfr(function(sd_delta, signal_share, rep) run_cell(sd_delta, signal_share, rep)) %>%
  group_by(sd_delta, signal_share) %>%
  summarise(var_delta = mean(var_delta), gap = mean(gap), .groups = "drop")
grid
```

```{r grid-plots, fig.width=9, fig.height=4}
p1 <- ggplot(grid, aes(var_delta, gap, colour = factor(signal_share))) +
  geom_point(size = 2) + geom_line() +
  labs(x = "Var(delta) (predictable signal)", y = "Empirical gap in MSE",
       colour = "Recoverable\nshare", title = "Gap grows with predictable variation") +
  theme_minimal(base_size = 12)
p2 <- ggplot(grid, aes(signal_share, gap, colour = factor(sd_delta))) +
  geom_point(size = 2) + geom_line() +
  labs(x = "Recoverable share of delta (proxy for rho^2)", y = "Empirical gap in MSE",
       colour = "sd(delta)", title = "Gap grows with how much the features recover") +
  theme_minimal(base_size = 12)
print(p1); print(p2)
dir.create("eda_outputs", showWarnings = FALSE)
ggsave("eda_outputs/fig_sim_gap_vardelta.pdf", p1, width = 6, height = 4)
ggsave("eda_outputs/fig_sim_gap_share.pdf",    p2, width = 6, height = 4)

# Wide grid for the thesis table (rows = sd_delta, cols = signal_share)
grid %>% select(sd_delta, signal_share, gap) %>%
  tidyr::pivot_wider(names_from = signal_share, values_from = gap) %>%
  knitr::kable(digits = 0, caption = "MSE gap by sd(delta) and recoverable share")
```

# 4. Interpretation

The empirical gap rises with both the strength of the predictable signal
(`Var(delta)`) and the share of it the features can recover (`signal_share`),
exactly the two terms in `Var(delta) * rho^2`. When either is near zero the tree
has nothing to add over the store mean. This reproduces, in a controlled
setting, the mechanism the empirical chapters test on Rossmann.

```{r session}
sessionInfo()
```
