---
title: "Rossmann Stage 1 — Tree-Based and Boosting Model Comparison"
author: "Kayra Işık"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  cache   = FALSE,
  fig.align = "center"
)
set.seed(42)
```

# Purpose

This file computes the **Stage 1** model comparison on the full Rossmann data
**without lag features**, training every model from scratch on the same
train/test split. It replaces the earlier setup in which the linear, CART,
bagging, random-forest, and from-scratch-GBM rows were carried over as stored
constants: here all nine models are fitted and evaluated in one reproducible
pass, so the whole table comes from a single run under the same pipeline.

Models compared:

- two naive baselines (store-mean, store-day-of-week),
- linear regression,
- a single CART tree,
- bagged trees,
- random forest,
- XGBoost,
- LightGBM.

# 1. Packages and configuration

```{r packages}
library(tidyverse)
library(lubridate)
library(rpart)
library(ranger)
library(xgboost)
library(lightgbm)

# ---- Paths: adjust if the CSVs live elsewhere -------------------------------
DATA_DIR  <- "."                       # folder containing train.csv and store.csv
TRAIN_CSV <- file.path(DATA_DIR, "train.csv")
STORE_CSV <- file.path(DATA_DIR, "store.csv")

# ---- Run configuration ------------------------------------------------------
DETECTED_CORES <- parallel::detectCores()
N_THREADS <- if (is.na(DETECTED_CORES)) 1L else max(1L, DETECTED_CORES - 1L)
TEST_DAYS <- 42        # last 42 days form the test set (June–July 2015)
VAL_FRAC  <- 0.15      # last 15% of the training dates used as a validation
                       # slice for boosting early stopping
SEED      <- 42
set.seed(SEED)
```

# 2. Data loading and preprocessing

```{r load-data}
train_raw <- read_csv(TRAIN_CSV, show_col_types = FALSE)
store_raw <- read_csv(STORE_CSV, show_col_types = FALSE)

train_raw <- train_raw %>% mutate(Date = as.Date(Date))
```

Closed days and zero-sales days are removed, the store table is joined, and a
compact set of calendar and store features is engineered.

```{r preprocess}
df <- train_raw %>%
  filter(Open == 1, Sales > 0) %>%
  left_join(store_raw, by = "Store") %>%
  mutate(
    Year       = year(Date),
    Month      = month(Date),
    Week       = isoweek(Date),
    DayOfMonth = mday(Date),
    Quarter    = quarter(Date),
    IsWeekend  = as.integer(DayOfWeek %in% c(6, 7)),
    StateHoliday = case_when(
      StateHoliday == "a" ~ 1L,
      StateHoliday == "b" ~ 2L,
      StateHoliday == "c" ~ 3L,
      TRUE                ~ 0L
    ),
    CompetitionOpenMonths = ifelse(
      !is.na(CompetitionOpenSinceYear) & !is.na(CompetitionOpenSinceMonth),
      (Year - CompetitionOpenSinceYear) * 12 + (Month - CompetitionOpenSinceMonth),
      NA_real_
    ),
    CompetitionOpenMonths = pmax(CompetitionOpenMonths, 0, na.rm = TRUE),
    StoreType  = as.integer(factor(StoreType,  levels = c("a", "b", "c", "d"))),
    Assortment = as.integer(factor(Assortment, levels = c("a", "b", "c")))
  ) %>%
  select(Store, Date, Sales, DayOfWeek, Promo, StateHoliday, SchoolHoliday,
         StoreType, Assortment, CompetitionDistance, Promo2,
         Year, Month, Week, DayOfMonth, Quarter, IsWeekend,
         CompetitionOpenMonths)

# Impute the few missing competition fields
med_comp_dist <- median(df$CompetitionDistance, na.rm = TRUE)
df <- df %>%
  mutate(
    CompetitionDistance   = replace_na(CompetitionDistance, med_comp_dist),
    CompetitionOpenMonths = replace_na(CompetitionOpenMonths, 0)
  ) %>%
  arrange(Date, Store)

cat("Rows after cleaning:", nrow(df), "\n")
cat("Stores:", n_distinct(df$Store), "\n")
cat("Date range:", as.character(min(df$Date)), "to",
    as.character(max(df$Date)), "\n")
```

```{r features}
# Stage 1 feature set (no lag / rolling features)
FEAT_S1 <- c("Store", "DayOfWeek", "Promo", "StateHoliday", "SchoolHoliday",
             "StoreType", "Assortment", "CompetitionDistance", "Promo2",
             "Year", "Month", "Week", "DayOfMonth", "Quarter", "IsWeekend",
             "CompetitionOpenMonths")
```

# 3. Train / test split

The split is temporal: the last `r TEST_DAYS` dates form the test set; all
earlier observations are used for training.

```{r split}
split_data <- function(data, feature_cols, test_days = TEST_DAYS) {
  data       <- data %>% arrange(Date, Store)
  all_dates  <- sort(unique(data$Date))
  cutoff     <- all_dates[length(all_dates) - test_days]

  train <- data %>% filter(Date <= cutoff)
  test  <- data %>% filter(Date >  cutoff)

  list(
    X_train = as.matrix(train[, feature_cols]),
    y_train = train$Sales,
    X_test  = as.matrix(test[, feature_cols]),
    y_test  = test$Sales,
    train   = train,
    test    = test
  )
}

split_s1 <- split_data(df, FEAT_S1)

cat("Train rows:", length(split_s1$y_train),
    "| Test rows:", length(split_s1$y_test), "\n")
cat("Test period:", as.character(min(split_s1$test$Date)), "to",
    as.character(max(split_s1$test$Date)), "\n")
```

# 4. Metrics

```{r metrics}
eval_metrics <- function(actual, pred) {
  valid <- !is.na(actual) & actual != 0
  tibble(
    RMSE = sqrt(mean((actual - pred)^2)),
    MAE  = mean(abs(actual - pred)),
    MAPE = mean(abs((actual[valid] - pred[valid]) / actual[valid])) * 100
  )
}

# Relative improvement (%) of a model over a reference RMSE
delta_rel <- function(ref_rmse, model_rmse) (ref_rmse - model_rmse) / ref_rmse * 100
```

# 5. Models

Each function takes the split and returns a one-row tibble of metrics.

## 5.1 Naive baselines

```{r m-naive}
run_store_mean <- function(split) {
  store_means <- split$train %>%
    group_by(Store) %>%
    summarise(m = mean(Sales), .groups = "drop")
  pred <- split$test %>%
    left_join(store_means, by = "Store") %>%
    mutate(m = replace_na(m, mean(split$train$Sales))) %>%
    pull(m)
  eval_metrics(split$y_test, pred)
}

run_store_dow <- function(split) {
  dow_means   <- split$train %>%
    group_by(Store, DayOfWeek) %>%
    summarise(m_dow = mean(Sales), .groups = "drop")
  store_means <- split$train %>%
    group_by(Store) %>%
    summarise(m_store = mean(Sales), .groups = "drop")
  global_mean <- mean(split$train$Sales)
  pred <- split$test %>%
    left_join(dow_means,   by = c("Store", "DayOfWeek")) %>%
    left_join(store_means, by = "Store") %>%
    mutate(pred = coalesce(m_dow, m_store, global_mean)) %>%
    pull(pred)
  eval_metrics(split$y_test, pred)
}
```

## 5.2 Linear regression

```{r m-linear}
run_linear <- function(split) {
  df_tr <- as.data.frame(split$X_train); df_tr$Sales <- split$y_train
  df_te <- as.data.frame(split$X_test)
  fit   <- lm(Sales ~ ., data = df_tr)
  pred  <- predict(fit, newdata = df_te)
  eval_metrics(split$y_test, pred)
}
```

## 5.3 CART (single regression tree)

```{r m-cart}
run_cart <- function(split) {
  df_tr <- as.data.frame(split$X_train); df_tr$Sales <- split$y_train
  df_te <- as.data.frame(split$X_test)
  fit   <- rpart(Sales ~ ., data = df_tr,
                 control = rpart.control(maxdepth = 20, cp = 1e-5, minsplit = 10))
  pred  <- predict(fit, newdata = df_te)
  eval_metrics(split$y_test, pred)
}
```

## 5.4 Bagging

```{r m-bagging}
run_bagging <- function(split, B = 200, min_node_size = 5) {
  df_tr <- as.data.frame(split$X_train); df_tr$Sales <- split$y_train
  df_te <- as.data.frame(split$X_test)
  p <- ncol(split$X_train)

  # Bagging is a random forest without feature subsampling: every split can use
  # all predictors (mtry = p), while each tree is grown on a bootstrap sample.
  fit <- ranger(
    Sales ~ ., data = df_tr,
    num.trees = B,
    mtry = p,
    min.node.size = min_node_size,
    replace = TRUE,
    sample.fraction = 1,
    num.threads = N_THREADS,
    seed = SEED,
    verbose = FALSE
  )
  pred <- predict(fit, data = df_te)$predictions
  eval_metrics(split$y_test, pred)
}
```

## 5.5 Random forest

```{r m-rf}
run_rf <- function(split, B = 200) {
  df_tr <- as.data.frame(split$X_train); df_tr$Sales <- split$y_train
  p     <- ncol(split$X_train)
  # mtry = p/3 is the standard regression default (ranger's own default is the
  # sqrt(p) classification rule, which over-subsamples here). Bagging is the same
  # model with mtry = p.
  fit   <- ranger(Sales ~ ., data = df_tr, num.trees = B, mtry = floor(p / 3),
                  num.threads = N_THREADS, seed = SEED, verbose = FALSE)
  pred  <- predict(fit, data = as.data.frame(split$X_test))$predictions
  eval_metrics(split$y_test, pred)
}
```

## 5.6 Gradient boosting from scratch

`rpart` trees fitted sequentially to the residuals, scaled by a learning rate.
This is the transparent reference implementation of boosting, not a tuned
library, but it is given enough capacity to be a fair competitor rather than an
under-powered straw man. Each round fits a depth-`depth` tree on a random
`row_frac` subsample of the rows (**stochastic** gradient boosting). The
subsampling roughly halves the per-tree cost *and* tends to improve
generalisation, so we can afford more rounds without the runtime exploding.
`B`, `lr`, `depth`, and `row_frac` trade runtime against accuracy; drop `B` to
100 for a quicker pass.

```{r m-gbm-scratch}
run_gbm_scratch <- function(split, B = 200, lr = 0.08, depth = 6,
                            row_frac = 0.5) {
  X_tr <- as.data.frame(split$X_train)
  X_te <- as.data.frame(split$X_test)
  y    <- split$y_train
  n    <- nrow(X_tr)

  F_tr <- rep(mean(y), n)
  F_te <- rep(mean(y), nrow(X_te))
  for (b in seq_len(B)) {
    idx  <- sample(n, floor(n * row_frac))          # stochastic row subsample
    df_b <- X_tr[idx, ]; df_b$resid <- (y - F_tr)[idx]
    fit  <- rpart(resid ~ ., data = df_b,
                  control = rpart.control(maxdepth = depth, cp = 0, minsplit = 20))
    F_tr <- F_tr + lr * predict(fit, newdata = X_tr)
    F_te <- F_te + lr * predict(fit, newdata = X_te)
  }
  eval_metrics(split$y_test, F_te)
}
```

## 5.7 XGBoost

Number of boosting rounds is chosen on a temporal validation slice (the last
`r scales::percent(VAL_FRAC)` of the training dates), then the model is refit on
the full training set.

```{r m-xgb}
temporal_val_index <- function(train_df, val_frac = VAL_FRAC) {
  dates  <- sort(unique(train_df$Date))
  cutoff <- dates[floor(length(dates) * (1 - val_frac))]
  which(train_df$Date > cutoff)            # rows used for validation
}

run_xgboost <- function(split, max_rounds = 1500, early_stop = 50) {
  X_tr <- split$X_train; y_tr <- split$y_train
  val_idx <- temporal_val_index(split$train)
  tr_idx  <- setdiff(seq_along(y_tr), val_idx)

  dtrain <- xgb.DMatrix(X_tr[tr_idx, ],  label = y_tr[tr_idx])
  dval   <- xgb.DMatrix(X_tr[val_idx, ], label = y_tr[val_idx])
  params <- list(objective = "reg:squarederror", eta = 0.10, max_depth = 6,
                 subsample = 0.8, colsample_bytree = 0.8,
                 nthread = N_THREADS, verbosity = 0)
  fit_es <- xgb.train(params, dtrain, nrounds = max_rounds,
                      watchlist = list(val = dval),
                      early_stopping_rounds = early_stop, verbose = 0)
  best_n <- fit_es$best_iteration

  fit <- xgb.train(params, xgb.DMatrix(X_tr, label = y_tr),
                   nrounds = best_n, verbose = 0)
  pred <- predict(fit, xgb.DMatrix(split$X_test))
  eval_metrics(split$y_test, pred)
}
```

## 5.8 LightGBM

```{r m-lgbm}
run_lightgbm <- function(split, max_rounds = 2000, early_stop = 50) {
  X_tr <- split$X_train; y_tr <- split$y_train
  val_idx <- temporal_val_index(split$train)
  tr_idx  <- setdiff(seq_along(y_tr), val_idx)

  dtrain <- lgb.Dataset(X_tr[tr_idx, ],  label = y_tr[tr_idx])
  dval   <- lgb.Dataset(X_tr[val_idx, ], label = y_tr[val_idx])
  params <- list(objective = "regression", metric = "rmse",
                 num_leaves = 63, learning_rate = 0.05,
                 bagging_fraction = 0.8, bagging_freq = 1,
                 feature_fraction = 0.8, num_threads = N_THREADS,
                 seed = SEED, verbosity = -1)
  fit_es <- lgb.train(params, dtrain, nrounds = max_rounds,
                      valids = list(val = dval),
                      early_stopping_rounds = early_stop, verbose = -1)
  best_n <- fit_es$best_iter

  fit <- lgb.train(params, lgb.Dataset(X_tr, label = y_tr),
                   nrounds = best_n, verbose = -1)
  pred <- predict(fit, split$X_test)
  eval_metrics(split$y_test, pred)
}
```

# 6. Run all models

```{r run-all}
models <- list(
  "Store-mean naive" = run_store_mean,
  "Store-DOW naive"  = run_store_dow,
  "Linear"           = run_linear,
  "CART"             = run_cart,
  "Bagging"          = run_bagging,
  "Random Forest"    = run_rf,
  "XGBoost"          = run_xgboost,
  "LightGBM"         = run_lightgbm
)

results <- map_dfr(names(models), function(nm) {
  set.seed(SEED)
  t0  <- proc.time()
  met <- models[[nm]](split_s1)
  cat(sprintf("%-18s done in %5.1f s\n", nm, (proc.time() - t0)["elapsed"]))
  met %>% mutate(Model = nm, .before = 1)
})
```

# 7. Results

RMSE is the primary ranking metric. `Delta_rel` is the improvement over the
store-mean baseline; `Delta_DOW` is the improvement over the stronger
store-day-of-week baseline.

```{r results-table}
rmse_mean <- results$RMSE[results$Model == "Store-mean naive"]
rmse_dow  <- results$RMSE[results$Model == "Store-DOW naive"]

results_tbl <- results %>%
  mutate(
    Delta_rel = delta_rel(rmse_mean, RMSE),
    Delta_DOW = delta_rel(rmse_dow,  RMSE)
  ) %>%
  arrange(RMSE) %>%
  mutate(across(c(RMSE, MAE), ~round(.x, 0)),
         across(c(MAPE, Delta_rel, Delta_DOW), ~round(.x, 1)))

results_tbl
```

```{r results-plot, fig.width=8, fig.height=5}
results_tbl %>%
  mutate(Model = fct_reorder(Model, RMSE)) %>%
  ggplot(aes(x = RMSE, y = Model)) +
  geom_col(fill = "#4C72B0", width = 0.65) +
  geom_text(aes(label = RMSE), hjust = -0.15, size = 3.4) +
  labs(title = "Stage 1 — RMSE by model (full Rossmann, no lag features)",
       x = "Test RMSE", y = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  theme_minimal(base_size = 12)
```

# 8. Session information

```{r session-info}
sessionInfo()
```
