---
title: "Rossmann Stages 2-4 — Subsetting, Lag Enrichment, Rolling Robustness"
author: "Kayra Işık"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
```

This file is **self-contained**: it defines the whole pipeline and runs Stages
2-4. Only `train.csv` and `store.csv` are needed (same folder). The pipeline
matches the Stage 1 file exactly.

# Pipeline (data, features, models)

```{r pipeline}
library(tidyverse); library(lubridate); library(zoo)
library(xgboost); library(lightgbm)

TRAIN_CSV <- "train.csv"; STORE_CSV <- "store.csv"
N_THREADS <- max(1, parallel::detectCores() - 1)
TEST_DAYS <- 42; VAL_FRAC <- 0.15; SEED <- 42; N_REPS <- 5
set.seed(SEED)

# ---- load + preprocess ----
train_raw <- read_csv(TRAIN_CSV, show_col_types = FALSE) %>% mutate(Date = as.Date(Date))
store_raw <- read_csv(STORE_CSV, show_col_types = FALSE)
df <- train_raw %>% filter(Open == 1, Sales > 0) %>% left_join(store_raw, by = "Store") %>%
  mutate(Year = year(Date), Month = month(Date), Week = isoweek(Date),
         DayOfMonth = mday(Date), Quarter = quarter(Date),
         IsWeekend = as.integer(DayOfWeek %in% c(6,7)),
         StateHoliday = case_when(StateHoliday=="a"~1L, StateHoliday=="b"~2L, StateHoliday=="c"~3L, TRUE~0L),
         CompetitionOpenMonths = ifelse(!is.na(CompetitionOpenSinceYear) & !is.na(CompetitionOpenSinceMonth),
           (Year-CompetitionOpenSinceYear)*12 + (Month-CompetitionOpenSinceMonth), NA_real_),
         CompetitionOpenMonths = pmax(CompetitionOpenMonths, 0, na.rm = TRUE),
         StoreType = as.integer(factor(StoreType, levels=c("a","b","c","d"))),
         Assortment = as.integer(factor(Assortment, levels=c("a","b","c")))) %>%
  select(Store, Date, Sales, DayOfWeek, Promo, StateHoliday, SchoolHoliday, StoreType,
         Assortment, CompetitionDistance, Promo2, Year, Month, Week, DayOfMonth,
         Quarter, IsWeekend, CompetitionOpenMonths)
med <- median(df$CompetitionDistance, na.rm = TRUE)
df <- df %>% mutate(CompetitionDistance = replace_na(CompetitionDistance, med),
                    CompetitionOpenMonths = replace_na(CompetitionOpenMonths, 0)) %>% arrange(Date, Store)

# ---- features ----
FEAT_S1 <- c("Store","DayOfWeek","Promo","StateHoliday","SchoolHoliday","StoreType","Assortment",
             "CompetitionDistance","Promo2","Year","Month","Week","DayOfMonth","Quarter","IsWeekend",
             "CompetitionOpenMonths")
LAG_COLS <- c("lag1","lag7","lag14","lag28"); ROLL_COLS <- c("roll7_mean","roll28_mean")
FEAT_S3 <- c(FEAT_S1, LAG_COLS, ROLL_COLS)
add_lag_features <- function(s) { s <- s %>% arrange(Date)
  s$lag1<-lag(s$Sales,1); s$lag7<-lag(s$Sales,7); s$lag14<-lag(s$Sales,14); s$lag28<-lag(s$Sales,28)
  s$roll7_mean<-lag(rollmeanr(s$Sales,7,fill=NA,na.rm=TRUE),1)
  s$roll28_mean<-lag(rollmeanr(s$Sales,28,fill=NA,na.rm=TRUE),1); s }

# ---- splits + metrics ----
split_data <- function(data, fc, td = TEST_DAYS) { data <- data %>% arrange(Date, Store)
  d <- sort(unique(data$Date)); cut <- d[length(d)-td]
  tr <- data %>% filter(Date<=cut); te <- data %>% filter(Date>cut)
  list(X_train=as.matrix(tr[,fc]), y_train=tr$Sales, X_test=as.matrix(te[,fc]), y_test=te$Sales, train=tr, test=te) }
split_window <- function(data, fc, end, td = TEST_DAYS, months = NULL) { data <- data %>% arrange(Date, Store)
  d <- sort(unique(data$Date)); d <- d[d<=end]; tds <- tail(d, td); ts <- min(tds)
  tr <- data %>% filter(Date<ts); if(!is.null(months)) tr <- tr %>% filter(Date >= ts %m-% months(months))
  te <- data %>% filter(Date %in% tds)
  list(X_train=as.matrix(tr[,fc]), y_train=tr$Sales, X_test=as.matrix(te[,fc]), y_test=te$Sales,
       train=tr, test=te, test_start=ts, test_end=max(tds)) }
eval_metrics <- function(a,p){ v <- !is.na(a) & a!=0
  tibble(RMSE=sqrt(mean((a-p)^2)), MAE=mean(abs(a-p)), MAPE=mean(abs((a[v]-p[v])/a[v]))*100) }
delta_rel <- function(ref,x) (ref-x)/ref*100

# ---- models ----
run_store_mean <- function(sp){ sm <- sp$train %>% group_by(Store) %>% summarise(m=mean(Sales),.groups="drop")
  p <- sp$test %>% left_join(sm,by="Store") %>% mutate(m=replace_na(m,mean(sp$train$Sales))) %>% pull(m)
  eval_metrics(sp$y_test,p) }
run_store_dow <- function(sp){ dm <- sp$train %>% group_by(Store,DayOfWeek) %>% summarise(md=mean(Sales),.groups="drop")
  sm <- sp$train %>% group_by(Store) %>% summarise(ms=mean(Sales),.groups="drop"); gm <- mean(sp$train$Sales)
  p <- sp$test %>% left_join(dm,by=c("Store","DayOfWeek")) %>% left_join(sm,by="Store") %>%
    mutate(p=coalesce(md,ms,gm)) %>% pull(p); eval_metrics(sp$y_test,p) }
val_idx <- function(tr,vf=VAL_FRAC){ d<-sort(unique(tr$Date)); which(tr$Date > d[floor(length(d)*(1-vf))]) }
run_xgboost <- function(sp, max_rounds=1500, es=50, return_model=FALSE){ X<-sp$X_train; y<-sp$y_train
  vi<-val_idx(sp$train); ti<-setdiff(seq_along(y),vi)
  pr<-list(objective="reg:squarederror", eta=0.10, max_depth=6, subsample=0.8,
           colsample_bytree=0.8, nthread=N_THREADS, verbosity=0)
  e<-xgb.train(pr, xgb.DMatrix(X[ti,],label=y[ti]), nrounds=max_rounds,
               watchlist=list(val=xgb.DMatrix(X[vi,],label=y[vi])), early_stopping_rounds=es, verbose=0)
  f<-xgb.train(pr, xgb.DMatrix(X,label=y), nrounds=e$best_iteration, verbose=0)
  p<-predict(f, xgb.DMatrix(sp$X_test))
  if(return_model) return(list(metrics=eval_metrics(sp$y_test,p), model=f, feat=colnames(X)))
  eval_metrics(sp$y_test,p) }
run_lightgbm <- function(sp, max_rounds=2000, es=50, return_model=FALSE){ X<-sp$X_train; y<-sp$y_train
  vi<-val_idx(sp$train); ti<-setdiff(seq_along(y),vi)
  pr<-list(objective="regression", metric="rmse", num_leaves=63, learning_rate=0.05,
           bagging_fraction=0.8, bagging_freq=1, feature_fraction=0.8, num_threads=N_THREADS, seed=SEED, verbosity=-1)
  e<-lgb.train(pr, lgb.Dataset(X[ti,],label=y[ti]), nrounds=max_rounds,
               valids=list(val=lgb.Dataset(X[vi,],label=y[vi])), early_stopping_rounds=es, verbose=-1)
  f<-lgb.train(pr, lgb.Dataset(X,label=y), nrounds=e$best_iter, verbose=-1)
  p<-predict(f, sp$X_test)
  if(return_model) return(list(metrics=eval_metrics(sp$y_test,p), model=f, feat=colnames(X)))
  eval_metrics(sp$y_test,p) }
run_panel <- function(data,label=""){ sp<-split_data(data,FEAT_S1)
  bm<-run_store_mean(sp)$RMSE; bd<-run_store_dow(sp)$RMSE; xg<-run_xgboost(sp)$RMSE; lg<-run_lightgbm(sp)$RMSE
  tibble(Condition=label, Naive_RMSE=bm, StoreDOW_RMSE=bd, XGB_RMSE=xg, LGBM_RMSE=lg,
         StoreDOW_delta=delta_rel(bm,bd), XGB_delta=delta_rel(bm,xg), LGBM_delta=delta_rel(bm,lg)) }

cat("Pipeline ready. Rows:", nrow(df), "Stores:", n_distinct(df$Store), "\n")
```

# Stage 2.1 — Number of stores

```{r s2-nstores, cache=TRUE}
all_stores <- unique(df$Store)
stage2_n <- map_dfr(c(50,200,500,length(all_stores)), function(N){
  reps <- if(N==length(all_stores)) 1 else N_REPS
  map_dfr(seq_len(reps), function(r){ set.seed(SEED+r)
    sub <- if(N==length(all_stores)) df else df %>% filter(Store %in% sample(all_stores,N))
    run_panel(sub, as.character(N)) }) }) %>%
  group_by(N=as.integer(Condition)) %>% summarise(across(where(is.numeric),mean),.groups="drop") %>%
  arrange(N) %>% mutate(across(where(is.numeric), ~round(.x,1)))
stage2_n
```

# Stage 2.2 — Training window length

```{r s2-window, cache=TRUE}
te <- max(df$Date)
stage2_window <- imap_dfr(list("6 months"=6,"12 months"=12,"18 months"=18,"24 months"=24,"Full"=NULL),
  function(mo,lab){ sp<-split_window(df,FEAT_S1,te,months=mo)
    bm<-run_store_mean(sp)$RMSE; bd<-run_store_dow(sp)$RMSE
    tibble(Window=lab, StoreDOW_delta=delta_rel(bm,bd),
           XGB_delta=delta_rel(bm,run_xgboost(sp)$RMSE), LGBM_delta=delta_rel(bm,run_lightgbm(sp)$RMSE)) }) %>%
  mutate(across(where(is.numeric), ~round(.x,1)))
stage2_window
```

# Stage 2.3 — Store-type composition

```{r s2-type, cache=TRUE}
stage2_type <- imap_dfr(list("A only"=function(d) filter(d,StoreType==1),
  "B only"=function(d) filter(d,StoreType==2), "A+B"=function(d) filter(d,StoreType %in% c(1,2)),
  "Full"=function(d) d), function(f,lab) run_panel(f(df),lab)) %>%
  mutate(across(where(is.numeric), ~round(.x,1)))
stage2_type
```

# Stage 3 — Lag feature enrichment (+ importance share)

```{r s3, cache=TRUE}
df_lags <- df %>% group_by(Store) %>% arrange(Date) %>% group_modify(~ add_lag_features(.x)) %>%
  ungroup() %>% filter(!is.na(lag28), !is.na(roll28_mean))
sp3 <- split_data(df_lags, FEAT_S3)
xgb3 <- run_xgboost(sp3, return_model=TRUE); lgb3 <- run_lightgbm(sp3, return_model=TRUE)
stage3 <- tibble(Model=c("Store-mean naive","Store-DOW naive","XGBoost","LightGBM"),
  RMSE=c(run_store_mean(sp3)$RMSE, run_store_dow(sp3)$RMSE, xgb3$metrics$RMSE, lgb3$metrics$RMSE))
rm3<-stage3$RMSE[1]; rd3<-stage3$RMSE[2]
stage3 <- stage3 %>% mutate(Delta_rel=delta_rel(rm3,RMSE), Delta_DOW=delta_rel(rd3,RMSE),
                            across(where(is.numeric), ~round(.x,1)))
stage3
lr_cols <- c(LAG_COLS, ROLL_COLS)
ix <- xgb.importance(model=xgb3$model, feature_names=xgb3$feat)
il <- lgb.importance(lgb3$model)
cat(sprintf("Lag/rolling gain-importance share — XGBoost: %.1f%%  LightGBM: %.1f%%\n",
  sum(ix$Gain[ix$Feature %in% lr_cols])/sum(ix$Gain)*100,
  sum(il$Gain[il$Feature %in% lr_cols])/sum(il$Gain)*100))
```

# Stage 4 — Rolling test-window robustness

```{r s4, cache=TRUE}
w1<-max(df$Date); w2<-w1-TEST_DAYS; w3<-w2-TEST_DAYS
stage4 <- imap_dfr(list("Window 1"=w1,"Window 2"=w2,"Window 3"=w3), function(end,lab){
  sp<-split_window(df,FEAT_S1,end)
  tibble(Window=lab, Test_period=paste(format(sp$test_start,"%d %b"),"-",format(sp$test_end,"%d %b %Y")),
         Store_mean=run_store_mean(sp)$RMSE, Store_DOW=run_store_dow(sp)$RMSE,
         XGBoost=run_xgboost(sp)$RMSE, LightGBM=run_lightgbm(sp)$RMSE) }) %>%
  mutate(across(where(is.numeric), ~round(.x)))
stage4
```

```{r session}
sessionInfo()
```
