---
title: "Characterising the Tree Advantage by Data Properties"
author: "Kayra Işık"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
set.seed(42)
```

# Purpose

The experiments vary operational knobs (number of stores, training length, store
type). This file reframes the question the way the supervisor asks for: it
describes the tree-based advantage in terms of **measurable data properties** of
each store's series, so the findings can be carried to other datasets with the
same or different characteristics.

For every store we compute a small set of series-level properties on the
training data, and the realised gain of a global LightGBM over the Store-DOW
baseline on the 42-day holdout. We then relate the gain to the properties by
regression. If the framework is right, the gain should rise with the share of
predictable within-store variation and with the strength of the weekly pattern,
and be roughly flat in the store's sales level and length.

# 1. Data

```{r packages}
library(tidyverse)
library(lubridate)
library(lightgbm)
TRAIN_CSV <- "train.csv"; STORE_CSV <- "store.csv"
N_THREADS <- max(1, parallel::detectCores() - 1); TEST_DAYS <- 42; SEED <- 42; set.seed(SEED)

train_raw <- read_csv(TRAIN_CSV, show_col_types = FALSE) %>% mutate(Date = as.Date(Date))
store_raw <- read_csv(STORE_CSV, show_col_types = FALSE)

# closed-day fraction per store (intermittency proxy) from the RAW data
closed <- train_raw %>% group_by(Store) %>%
  summarise(closed_frac = mean(Open == 0), .groups = "drop")

df <- train_raw %>% filter(Open == 1, Sales > 0) %>% left_join(store_raw, by = "Store") %>%
  mutate(Year=year(Date), Month=month(Date), Week=isoweek(Date), DayOfMonth=mday(Date),
         Quarter=quarter(Date), IsWeekend=as.integer(DayOfWeek %in% c(6,7)),
         StateHoliday=case_when(StateHoliday=="a"~1L,StateHoliday=="b"~2L,StateHoliday=="c"~3L,TRUE~0L),
         CompetitionOpenMonths=ifelse(!is.na(CompetitionOpenSinceYear)&!is.na(CompetitionOpenSinceMonth),
           (Year-CompetitionOpenSinceYear)*12+(Month-CompetitionOpenSinceMonth),NA_real_),
         CompetitionOpenMonths=pmax(CompetitionOpenMonths,0,na.rm=TRUE),
         StoreType=as.integer(factor(StoreType,levels=c("a","b","c","d"))),
         Assortment=as.integer(factor(Assortment,levels=c("a","b","c")))) %>%
  select(Store,Date,Sales,DayOfWeek,Promo,StateHoliday,SchoolHoliday,StoreType,Assortment,
         CompetitionDistance,Promo2,Year,Month,Week,DayOfMonth,Quarter,IsWeekend,CompetitionOpenMonths)
med <- median(df$CompetitionDistance,na.rm=TRUE)
df <- df %>% mutate(CompetitionDistance=replace_na(CompetitionDistance,med),
                    CompetitionOpenMonths=replace_na(CompetitionOpenMonths,0)) %>% arrange(Date,Store)

d <- sort(unique(df$Date)); cutoff <- d[length(d)-TEST_DAYS]
train <- df %>% filter(Date <= cutoff); test <- df %>% filter(Date > cutoff)
FEAT <- c("Store","DayOfWeek","Promo","StateHoliday","SchoolHoliday","StoreType","Assortment",
          "CompetitionDistance","Promo2","Year","Month","Week","DayOfMonth","Quarter","IsWeekend","CompetitionOpenMonths")
cat("Stores:", n_distinct(df$Store), "| train rows:", nrow(train), "\n")
```

# 2. Per-store data properties (training data only)

```{r properties}
props <- train %>% group_by(Store) %>% group_modify(function(g, key) {
  n <- nrow(g); v <- var(g$Sales)
  dow_mean   <- ave(g$Sales, g$DayOfWeek)
  cell_mean  <- ave(g$Sales, g$DayOfWeek, g$Promo)
  trend_fit  <- lm(Sales ~ seq_len(n), data = g)
  tibble(
    mean_sales    = mean(g$Sales),
    cv            = sd(g$Sales) / mean(g$Sales),
    dow_strength  = var(dow_mean)  / v,                 # weekly predictability
    pred_share    = var(cell_mean) / v,                 # Var(delta) proxy (DOW x Promo)
    trend_strength= summary(trend_fit)$r.squared,
    promo_frac    = mean(g$Promo),
    length        = n
  )
}) %>% ungroup() %>% left_join(closed, by = "Store")
summary(props %>% select(-Store))
```

# 3. Per-store realised gain (global LightGBM vs Store-DOW)

```{r gain, cache=TRUE}
# Store-DOW baseline predictions on the holdout
dowm <- train %>% group_by(Store, DayOfWeek) %>% summarise(md = mean(Sales), .groups="drop")
sm   <- train %>% group_by(Store) %>% summarise(ms = mean(Sales), .groups="drop")
gm   <- mean(train$Sales)
test_pred <- test %>% left_join(dowm, by=c("Store","DayOfWeek")) %>% left_join(sm, by="Store") %>%
  mutate(pred_dow = coalesce(md, ms, gm))

# one global LightGBM (no lags), predictions on the holdout
mod <- lgb.train(list(objective="regression", metric="rmse", num_leaves=63,
                      learning_rate=0.05, bagging_fraction=0.8, bagging_freq=1,
                      feature_fraction=0.8, num_threads=N_THREADS, seed=SEED, verbosity=-1),
                 lgb.Dataset(as.matrix(train[,FEAT]), label=train$Sales),
                 nrounds=400, verbose=-1)
test_pred$pred_lgbm <- predict(mod, as.matrix(test[,FEAT]))

gain <- test_pred %>% group_by(Store) %>%
  summarise(rmse_dow  = sqrt(mean((Sales-pred_dow)^2)),
            rmse_lgbm = sqrt(mean((Sales-pred_lgbm)^2)), .groups="drop") %>%
  mutate(gain_pct = 100 * (rmse_dow - rmse_lgbm) / rmse_dow)
summary(gain$gain_pct)
```

# 4. Relating the gain to data properties

```{r regression}
dat <- inner_join(props, gain, by = "Store")

# standardise predictors so coefficients are comparable
preds <- c("pred_share","dow_strength","cv","trend_strength","promo_frac",
           "closed_frac","mean_sales","length")
datz <- dat %>% mutate(across(all_of(preds), ~ as.numeric(scale(.x))))
fit  <- lm(reformulate(preds, "gain_pct"), data = datz)
summary(fit)
```

```{r coef-plot, fig.width=7, fig.height=4}
dir.create("eda_outputs", showWarnings = FALSE)
p_coef <- broom::tidy(fit) %>% filter(term != "(Intercept)") %>%
  mutate(term = fct_reorder(term, estimate)) %>%
  ggplot(aes(estimate, term)) +
  geom_col(fill = "#4C72B0") +
  geom_vline(xintercept = 0, colour = "grey40") +
  labs(x = "Std. coefficient on per-store gain (% over Store-DOW)", y = NULL,
       title = "Which data properties predict the tree advantage?") +
  theme_minimal(base_size = 12)
p_coef
ggsave("eda_outputs/fig_gain_vs_properties.pdf", p_coef, width = 7, height = 4)
```

```{r scatter, fig.width=9, fig.height=4}
p1 <- ggplot(dat, aes(pred_share, gain_pct)) + geom_point(alpha=.3, size=1) +
  geom_smooth(method="lm", se=TRUE) +
  labs(x="Predictable share Var(delta)/Var(y)", y="Gain over Store-DOW (%)") +
  theme_minimal(base_size=11)
p2 <- ggplot(dat, aes(cv, gain_pct)) + geom_point(alpha=.3, size=1) +
  geom_smooth(method="lm", se=TRUE) +
  labs(x="Coefficient of variation", y=NULL) + theme_minimal(base_size=11)
print(p1); print(p2)
```

# 5. Interpretation

The regression says which **data properties**, not which operational settings,
drive the tree advantage. A store gains more from the boosted model when more of
its within-store variance is predictable from the available features
(`pred_share`, `dow_strength`), which is exactly the mechanism in
Equation \eqref{eq:gap}. Properties like the sales level (`mean_sales`) and the
length of the series carry little once predictability is accounted for. Because
these properties can be measured on any retail series, the mapping from
properties to expected gain is what would transfer to a different dataset, which
is the generalisable form of the result.

```{r session}
sessionInfo()
```
